\normalfont\spacedallcaps{Hazardous asteroids forecast via Markov random fields}

### 2021-11-30 Manuscript/manuscript.tex
PUB

%*********************************************************************************
% Personal
%*********************************************************************************
\usepackage[font=itshape]{quoting}			% fancy quotation packages. [font=small] old option
\usepackage[english]{varioref}		% complete reference package
\usepackage{hyperref}
\usepackage{url}
\usepackage[intoc, english, noprefix]{nomencl}	%for list of symbols
\usepackage[normalem]{ulem}
\usepackage{chemfig} %for chemical formulas
\usepackage{eurosym}			% euro symbol
\usepackage{epigraph}
\usepackage{calligra}
\usepackage{soul}
\usepackage[utf8]{inputenc}

@@

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the theoretical approach, in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., a symmetry, the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynmanlectures}), the machine learning algorithms start from very general assumptions/expressions and then their parameters, are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecast with a reduced effort. However, the toll that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of conditional independences/dependencies, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its feature was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest (RF), Support Vector Machines (SVM), and Logistic Regression and Quadric Discriminant analysis (QDA). The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of conditional dependencies correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism by which a forecast is preferred to another one may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contain the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. This paper is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

In this section, I am going to review the theoretical concepts
that underlie the probabilistic methods here used: these will be
exposed following the approaches of Murphy \cite{murphy2012machine} , Koller
et al. \citep{koller2009probabilistic}, Højsgaardand et al. \citep{hojsgaard2012graphical} and Russel et al. \citep{russell2010artificial}. Furthermore, a rapid overview of the main concepts of information theory will also be provided following the Cover \citep{cover2006elements}, and MacKay \citep{mackay2003information} approaches. This is because some of its concepts in the preliminary analysis of the dataset were used. Finally, I concluded this chapter with a rapid overview of algorithm interpretability, which, together with the accuracy and the
confusion matrix, is necessary for the assessment of a ML performances. On the other side, the concepts related to the celestial
mechanics here used will be described, following the Murray
approach \cite{murray1999solar} into the Appendix A.

@@

Lets start by supposing that we would represent compactly a joint distribution such as \cite{murphy2012machine}:
\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}
that can represent for instance words in a documents or pixels of an image.  Firstly we know that using the chain rule,  we can decompose it, into the following form \cite{murphy2012machine}:
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
Where V is the number of variables and 1:V stands for ${1,2,...,V}$.  This decomposition makes explicit the conditional probability tables, or in other terms, the transition probability tensors \cite{wu2017markov}.  As one can point out, the number of parameters is cumbersome as the number of variables grows: indeed, the number of parameter required scales as $\mathcal{O}(K^{V})$.  Such a challenging problem can be attacked by considering the concept of conditional independence. This is defined as\cite{murphy2012machine}:

@@

In the first term $\textbf{y}$ is fixed to its observed values while in the second it is free. Such expression can be recasted in to a more explicative form \cite{murphy2012machine}:

@@

Therfore at the optimum we will have \cite{murphy2012machine}:

@@

\section{Information theory}
Given an ensemble of random variables, we can quantify the amount of information that one variable contains: such quantity is called mutual information and is a crucial concept within information theory. This approach, which was implemented by Claude Shannon decades before the probabilistic modelling, represents a complementary way by which one can attack the problem of conditional dependence between random variables. Furthermore, as shown in Fig. \ref{Information_theory_connections}, this theory provides a formidable contribution to different scientific fields. Here I would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allow defining the concept of mutual information properly. The founding concept of information theory is entropy. This quantity expresses the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficiently the second one describes the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated with another cannot have a degree of uncertainty lower to the quantity it aims to describe. On these bases, we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
 As for the joint distribution also in this case, we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
The concepts here reviewed, and the relations that interconnect them can be represented as shown in Fig. \ref{Entropy_MI}. Finally, we would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chain, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise, we would have a clear violation of the second principle of thermodynamics (see for instance, the Maxwell's demon \cite{feynman2018feynman}).

@@

Besides the accuracy and other features connected with the confusion matrix, another critical feature is algorithm interpretability. This feature represents how much the algorithm explains its predictions and, in general, its mechanics. In other words, how much the algorithm is not a black box. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g. right to explanation)}, in the author opinion, results provided by such an algorithm cannot be considered scientific since the scientific method also requires to explain and not simply almost correct forecast. Following the Tarski et al. \cite{tarski1953undecidable} argument, which is based on mathematical logic, it can be said that the formal theory T can be translated into S if and only if S can prove the theorem of T in its language. On the other side, we would also know that an algorithm, as an explanation, is complete: we expect that it will also be able to make correct forecasts for all available data. However, as shown by \cite{doshi2017towards}, and by \cite{gilpin2018explaining} the interpretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal systems as algorithms or scientific theories. For instance, let us consider the fall of object: at first glance, one can consider the only the items on Earth, in this case, the acceleration is constant, and it is a straightforward theory; but as one moves to consider also the interaction within planets a much more complicate law should be considered (in which the previous case is a particular one). The first one is fully explainable but poorly complete; the second one is more difficult to explain but much more complete. Therefore, in principle, the evaluation of algorithm performances should be considered not only into one dimension, the accuracy but also by considering its interpretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into a more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research centre. Among the 40 features that were present in the dataset,  the following cases were excluded: the redundant features (e.g. the distance quantity evaluated in miles instead of kilometres), the features connected only to the other name of the asteroid, the features connected to the name of the orbit and the one connected with the orbiting planet ( since for all it was the Earth). Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A since their description cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only 16 \% of the asteroids were hazardous, I considered reducing the number of the non-hazardous to improve the portion of the hazardous ones. For this purpose, I constructed a database in which the proportion hazardous/not hazardous was 1:5: thus, all the hazardous ones were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle, the proportion should be much less unbalanced, however, the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here is used as a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

@@

This chapter is organized in the following way: first, we will report the results concerning the preliminary analysis performed on the dataset. The factor analysis of mixed data and the mutual information analysis of the continuous variables of asteroids vs their hazard will be reported. Then it will follow the analysis of the dataset performed with the probabilistic methods. After a preliminary analysis on the continuous variables, the mixed interaction model and the minforest model obtained for the whole dataset will be presented and discussed. Finally, the probabilistic models previously obtained will be compared with the outputs and the performances of four machine learning algorithms (Random Forest, Support vector machines, Quadratic Discriminant Analysis and Logistic regression).

@@

The first inspection that was performed on the dataset was related to the density distributions of a selection of continuous features that are known, from the celestial mechanics, to be important for the prediction of the asteroids dangerousness. These are reported in Fig. \ref{Density_relevant}. Next we moved on a more systematic analysis with FADM and mutual information. In Fig. \ref{FADM},\ref{FAMD_Quantitative variables} and \ref{FAMD_Individuals_(c)} are reported the main achivements of FADM (performed with FactoMineR package \cite{le2008factominer}): in particular from the correlation circle in Fig. \ref{FAMD_Quantitative variables} the different correlations that are given by the celestial mechanics laws can be recognized: for instance, see that there is strong correlation of the mean motion with the semi-major axis. This is due to the Kepler law discussed in Appendix A. Furthermore we see that that the mean motion is correctly almost independent with respect to the the diameter (max or min) of  the asteroid. As explained in Appendix A this is an another result of Newton gravitation theory for a two body interaction: the mean motion of an asteroid, as long as the the mass of it is much lower with respect to the sun, is independent from its mass. On the other side the fact that relative velocity is correlated with the diameter is a spurios correlation. At this point one can ask why the mean motion and the relative velocity are orthogonal: this point will be clarified from a theoretical point of view in appendix A and also with graphical models. Briefly the motion of an object on an ellipse as seen from a focus is not uniform: this is faster as the two bodies approaches. Beside this inspection a further analysis based on the concept of mutual information (summarized in the previous chapter) was performed: its result is reported in Fig. \ref{Mutual_information}. This figure summarize the ranking of the features considered in the dataset for the dangerousness classification of the asteroids according to the following expression \cite{kratzer2018varrank}

@@

following the Peng. et al approach \cite{peng2005feature}. We see that the first place in the mutual information ranking, looking to the diagonal element, is taken by minimum orbit intersection: this is correct since this is one parameter used by NASA for deciding if an asteroid is hazardous or not (see Appendix A). The second place is occupied by Epoch date close approach. The third place is the eccentricity value: this parameter is entangled with the minimum orbit intersection and thus it is reasonable that it is important. Then we have two parameter that are related to the dimension of the asteroid: this is meaningful since if the asteroid has a too reduced volume it will be destroyed by the Earth atmosphere. On the other side the other parameters seems to have a too low MI for being interesting in this preliminary analysis. The results obtained for FAMD and MI will provide a useful path-guide, together with the celestial mechanics theory, for the interpretation of the results obtained in the following section

>>>>

%*********************************************************************************
% Personal
%*********************************************************************************
\usepackage[font=itshape]{quoting}			% fancy quotation packages. [font=small] old option
\usepackage[english]{varioref}		% complete reference package
\usepackage{hyperref}
\usepackage{url}
\usepackage[intoc, english, noprefix]{nomencl}	%for list of symbols
\usepackage[normalem]{ulem}
\usepackage{chemfig} %for chemical formulas
\usepackage{eurosym}			% euro symbol
\usepackage{epigraph}
\usepackage{calligra}
\usepackage{soul}
\usepackage[utf8]{inputenc}
\usepackage{csquotes}

@@

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the theoretical approach, in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., a symmetry, the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynmanlectures}), the machine learning algorithms start from very general assumptions/expressions and then their parameters are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecast with a reduced effort. However, the toll that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of conditional independences/dependencies, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its features was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest (RF), Support Vector Machines (SVM), and Logistic Regression and Quadric Discriminant analysis (QDA). The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of conditional dependencies correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism,  by which a forecast is preferred to another one, may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contains the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. This paper is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

In this section, I am going to review the theoretical concepts
that underlie the probabilistic methods here used: these will be
exposed following the approaches of Murphy \cite{murphy2012machine} , Koller
et al. \citep{koller2009probabilistic}, Højsgaardand et al. \citep{hojsgaard2012graphical} and Russel et al. \citep{russell2010artificial}. Furthermore, a rapid overview of the main concepts of information theory will also be provided following the Cover \citep{cover2006elements}, and MacKay \citep{mackay2003information} approaches. This is because some of its quantities in the preliminary analysis of the dataset were used. Finally, I concluded this chapter with a rapid overview of algorithm interpretability, which, together with the accuracy and the
confusion matrix, is necessary for the assessment of a ML performances. On the other side, the concepts related to the celestial
mechanics here used will be described, following the Murray
approach \cite{murray1999solar} into the Appendix A.

@@

Lets start by supposing that we would represent compactly a joint distribution such as \cite{murphy2012machine}:
\begin{equation}
p(x_{1},x_{2},...,x_{n})
\end{equation}
that can represent for instance words in a documents or pixels of an image.  Firstly we know that using the chain rule,  we can decompose it, into the following form \cite{murphy2012machine}:
\begin{equation}
p(x_{1:V})=p(x_{1})p(x_{2}|x_{1})p(x_{3}|x_{2},x_{1})...p(x_{V}|x_{1:V-1})
\end{equation}
Where V is the number of variables and $1:V$ stands for ${1,2,...,V}$.  This decomposition makes explicit the conditional probability tables, or in other terms, the transition probability tensors \cite{wu2017markov}.  As one can point out, the number of parameters is cumbersome as the number of variables grows: indeed, the number of parameter required scales as $\mathcal{O}(K^{V})$.  Such a challenging problem can be attacked by considering the concept of conditional independence. This is defined as\cite{murphy2012machine}:

@@

In the first term $\textbf{y}$ is fixed to its observed values while in the second it is free. Such expression can be recast in to a more explicative form \cite{murphy2012machine}:

@@

Therefore at the optimum we will have \cite{murphy2012machine}:

@@

\section{Information theory}
Given an ensemble of random variables, we can quantify the amount of information that one variable contains: such quantity is called mutual information and is a crucial concept within information theory. This approach, which was implemented by Claude Shannon decades before the probabilistic modelling, represents a complementary way by which one can attack the problem of conditional dependence between random variables. Furthermore, as shown in Fig. \ref{Information_theory_connections}, this theory provides a formidable contribution to different scientific fields. Here I would provide some basic concepts of this theory, following the Cover \cite{cover2006elements} and MacKay \cite{mackay2003information} approaches, that allow defining the concept of mutual information properly. The founding concept of information theory is entropy. This quantity expresses the uncertainty of a random variable. Given a random variable $X$ with alphabet (the accessible states) $\cite{cover2006elements}$ and probability mass function $p(x)=Pr\left\lbrace X=x \rbrace\; x \in \chi$ we define the entropy of $X$ as $H(X)=-\sum_{x \in X} p(x)log p(x)$ where the logarithm has to be considered with basis 2. In analogous way the joint entropy of two random variables $(X,Y)$ with a joint distribution p(x,y) is defined as \cite{cover2006elements}:
\begin{equation}
H(X,Y)=-\sum_{x\in \chi}\sum_{y\in \mathcal{Y}}p(x,y)\log p(x,y)
\end{equation}
Furthermore, we can define also the conditional entropy as \cite{cover2006elements}:
\begin{equation}
\begin{split}
H(X|Y)=& \sum_{x\in \mathcal{X} }p(x)H(Y|X=x)\\
=& -\sum_{x\in \mathcal{X}}p(x)\sum_{y\in \mathcal{Y}}p(x,y)\log p(y|x) \\
=& -E\log p(Y|X)
\end{split}
\end{equation}
The joint entropy and the conditional entropy are related by the chain rule \cite{cover2006elements}:
\begin{equation}
H(X,Y)=H(X)+H(Y|X)
\end{equation}
Such rule can be extended to to the following from \cite{cover2006elements}:
\begin{equation}
H(X,Y|Z)=H(X|Z)+H(Y|X,Z)
\end{equation}
Given a distribution q and another distribution p, one can quantify how inefficiently the second one describes the first one using the concept of relative entropy or Kullback-Leibler distance \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)=\sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}
As stated by the Gibbs inequality \cite{cover2006elements,mackay2003information}:
\begin{equation}
D(p||q)\geq 0
\end{equation}
this quantity can not be negative: the entropy of a random variable associated with another cannot have a degree of uncertainty lower to the quantity it aims to describe. On these bases, we are now ready to introduce the concept of mutual information. This is defined as  \cite{cover2006elements}:
\begin{equation}
\begin{split}
I(X;Y)&=\sum\sump(x,y)\log\dfrac{p(x,y)}{p(x)p(y)}=\\
&=D(p(x,y)||p(x)p(y)) \\
&=H(X)-H(X|Y)=H(Y)-H(Y|X)
\end{split}
\end{equation}
 As for the joint distribution also in this case, we have a chain rule  \cite{cover2006elements}:
\begin{equation}
I(X_{1},X_{2},...,X_{n};Y)=\sum^{n}_{i=1}I(X_{i};Y|X_{i-1},X_{i-2},...,X_{1})
\end{equation}
The concepts here reviewed, and the relations that interconnect them can be represented as shown in Fig. \ref{Entropy_MI}. Finally, I would report the data process inequality theorem that connects the information theory with the Markov chain: if we have a Markov chain, $X\rightarrow Y \rightarrow Z$  then $I(X;Y)\geq I(X;Z) $. As for the Gibbs inequality, the underlying idea is that no clever manipulation of the data can improve the inference that can be made on them \cite{cover2006elements,mackay2003information}. Otherwise, we would have a clear violation of the second principle of thermodynamics (see for instance, the Maxwell's demon \cite{feynman2018feynman}).

@@

Besides the accuracy and other features connected with the confusion matrix, another critical feature is algorithm interpretability. This concept represents how much the algorithm explains its predictions and, in general, its mechanics. In other words, how much the algorithm is not a black box. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g. right to explanation)}, in the author opinion, results provided by such an algorithm cannot be considered scientific since the scientific method also requires to explain and not simply almost correct forecast. Following the Tarski et al. \cite{tarski1953undecidable} argument, which is based on mathematical logic, it can be said that the formal theory T can be translated into S if and only if S can prove the theorem of T in its language. On the other side, we would also know that an algorithm, as an explanation, is complete: we expect that it will also be able to make correct forecasts for all available data. However, as shown by \cite{doshi2017towards}, and by \cite{gilpin2018explaining} the interpretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal systems as algorithms or scientific theories. For instance, let us consider the fall of object: at first glance, one can consider only the items on Earth. In this case, the acceleration is constant, and a straightforward theory will be obtained.  As one moves to consider also the interaction within planets and stars a much more complicate law should be considered (in which the previous case is a particular one). The first one is fully explainable but poorly complete; the second one is more difficult to explain but much more complete.  This is something that was elegantly stated by A.Einstein \cite{physics-reality}:
\begin{displayquote}
\textit{Science uses the totality of the primary concepts, i.e., concepts directly connected with sense experiences, and propositions connecting them. In its first stage of development, science does not contain anything else. Our everyday thinking is satisfied on the whole with this level. Such a state of affairs cannot, however, satisfy a spirit which is really scientifically minded; because the totality of concepts and relations obtained in this manner is utterly lacking in logical unity. In order to supplement this deficiency, one invents a system poorer in concepts and relations, a system retaining the primary concepts and relations of the “first layer” as logically derived concepts and relations. This new “secondary system” pays for its higher logical unity by having elementary concepts (concepts of the second layer), which are no longer directly connected with complexes of sense experiences. Further striving for logical unity brings us to a tertiary system, still poorer in concepts and relations, for the deduction of the concepts and relations of the secondary (and so indirectly of the primary) layer. Thus the story goes on until we have arrived at a system of the greatest conceivable unity, and of the greatest poverty of concepts of the logical foundations, which is still compatible with the observations made by our senses. We do not know whether or not this ambition will ever result in a definitive system. If one is asked for his opinion, he is inclined to answer no. While wrestling with the problems, however, one will never give up hope that this greatest of all aims can really be attained to a very high degree [...]The essential thing is the aim to represent the multitude of concepts and propositions, close to experience, as propositions, logically deduced from a basis, as narrow as possible, of fundamental concepts and fundamental relations which themselves can be chosen freely (axioms). The liberty of choice, however, is of a special kind; it is not in any way similar to the liberty of a writer of fiction. Rather, it is similar to that of a man engaged in solving a well-designed word puzzle. He may, it is true, propose any word as the solution; but, there is only one word which really solves the puzzle in all its parts. It is a matter of faith that nature– as she is perceptible to our five senses– takes the character of such a well-formulated puzzle. The successes reaped up to now by science do, it is true, give a certain encouragement for this faith.}
\end{displayquote}
Therefore, in principle, the evaluation of algorithm performances should be considered not only into one dimension, the accuracy but also by considering its interpretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into a more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research centre. Among the 40 features that were present in the dataset,  the following cases were excluded: the redundant features (e.g. the distance quantity evaluated in miles instead of kilometres), the features connected only to the other name of the asteroid, the features connected to the name of the orbit and the one connected with the orbiting planet ( since for all it was the Earth). Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A since tit cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only 16 \% of the asteroids were hazardous, I considered reducing the number of the non-hazardous to improve the portion of the hazardous ones. For this purpose, I constructed a database in which the proportion hazardous/not hazardous was 1:5: thus, all the hazardous ones were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle, the proportion should be much less unbalanced, however, the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here is used as a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

@@

This chapter is organized in the following way: first, I will report the results concerning the preliminary analysis performed on the dataset. The factor analysis of mixed data and the mutual information analysis of the continuous variables of asteroids vs their hazard will be reported. Then it will follows the analysis of the dataset performed with the probabilistic methods. After a preliminary analysis on the continuous variables, the mixed interaction model and the minforest model obtained for the whole dataset will be presented and discussed. Finally, the probabilistic models previously obtained will be compared with the outputs and the performances of four machine learning algorithms (Random Forest, Support vector machines, Quadratic Discriminant Analysis and Logistic regression).

@@

The first inspection that was performed on the dataset was related to the density distributions of a selection of continuous features that are known, from the celestial mechanics, to be important for the prediction of the asteroids dangerousness. These are reported in Fig. \ref{Density_relevant}. Next we moved on a more systematic analysis with FADM and mutual information. In Fig. \ref{FADM},\ref{FAMD_Quantitative variables} and \ref{FAMD_Individuals_(c)} are reported the main achivements of FADM (performed with FactoMineR package \cite{le2008factominer}): in particular from the correlation circle in Fig. \ref{FAMD_Quantitative variables} the different correlations that are given by the celestial mechanics laws can be recognized: for instance, see that there is strong anti-correlation of the mean motion with the semi-major axis. This is due to the Kepler law discussed in Appendix A. Furthermore we see that that the mean motion is correctly almost independent with respect to the the diameter (max or min) of  the asteroid. As explained in Appendix A this is an another result of Newton gravitation theory for a two body interaction: the mean motion of an asteroid, as long as the the mass of it is much lower with respect to the sun, is independent from its mass. On the other side the fact that relative velocity is correlated with the diameter is a spurios correlation. At this point one can ask why the mean motion and the relative velocity are orthogonal: this point will be clarified from a theoretical point of view in appendix A and also with graphical models. Briefly the motion of an object on an ellipse as seen from a focus is not uniform: this is faster as the two bodies approaches. Beside this inspection a further analysis based on the concept of mutual information (summarized in the previous chapter) was performed: its result is reported in Fig. \ref{Mutual_information}. This figure summarize the ranking of the features considered in the dataset for the dangerousness classification of the asteroids according to the following expression \cite{kratzer2018varrank}

@@

following the Peng. et al approach \cite{peng2005feature}. It can be seen that the first place in the mutual information ranking, looking to the diagonal element, is taken by minimum orbit intersection: this is correct since this is one parameter used by NASA for deciding if an asteroid is hazardous or not (see Appendix A). The second place is occupied by Epoch date close approach. The third place is the eccentricity value: this parameter is entangled with the minimum orbit intersection and thus it is reasonable that it is important. Then two parameters that are related to the dimension of the asteroid are present: this is meaningful since if the asteroid has a too reduced volume it will be destroyed by the Earth atmosphere. On the other hand the other parameters seems to have a too low MI for being interesting in this preliminary analysis. The results obtained for FAMD and MI will provide a useful path-guide, together with the celestial mechanics theory, for the interpretation of the results obtained in the following section.
